Pipelines: a fundamental building block for efficient hardware

### 2018-03-05 Documents/Tutorials/Aa/Aa.tex
Updated the tutorial

\documentclass{beamer}
\usepackage{epsfig}
\title{An introduction to the {\bf Aa} language}
\author{Madhav P. Desai}
\begin{document}
\maketitle

@@

\frame[containsverbatim]{\frametitle{{\bf Aa} Objects}
\begin{itemize}
\item Storage objects:
\begin{verbatim}
$storage A: $array [1024] $of $uint<32>
\end{verbatim}
\item Constant objects:
\begin{verbatim}
$constant A: $uint<4> := _b0011
$constant B: $uint<32> := _hf0f0f0f0
$constant C: $float<8,23> := _f2.3465e+0
\end{verbatim}
\item Pipe objects:
\begin{verbatim}
$pipe A: $uint<32> $depth 4
$lifo $pipe B: $uint<8> $depth 8
\end{verbatim}
\item Implicit Variables:  these are defined by
statements:
\begin{verbatim}
a := (A + B)
\end{verbatim}
They are also called {\em static-single-assignment}
or SSA variables.
\end{itemize}
}

@@

@@

\frame[containsverbatim]{\frametitle{{\bf Aa} Do-While Statements}
These are not atomic, and can occur only inside a branch-block.
\begin{verbatim}
$do
   $merge $entry $loopback
        $phi I := ($bitcast ($uint<32>) 0) 
                      $on $entry
                   NI $on $loopback
   $endmerge
   a[I] := (b[I] + c[I])
   NI   := (I+1)
$while (I < 16)
\end{verbatim}
Control-flow:  sequential, controlled by the condition check.  The
places \$entry and \$loopback are implicitly defined.  When control
enters the do-while, the token gets placed in \$entry and when
control loops-back from the condition check, the token gets placed
in \$loopback.
}

\frame[containsverbatim]{\frametitle{Getting to hardware starting from an {\bf Aa} Program}
}

>>>>

\documentclass{beamer}
\usepackage{epsfig}
\title{An introduction to the {\bf Aa} language}
\author{Madhav P. Desai}
\date{March 8, 2018}
\begin{document}
\maketitle

@@

\frame[containsverbatim]{\frametitle{{\bf Aa} Objects}
\begin{itemize}
\item Storage objects:
\begin{verbatim}
$storage A: $array [1024] $of $uint<32>
\end{verbatim}
\item Constant objects:
\begin{verbatim}
$constant A: $uint<4> := _b0011
$constant B: $uint<32> := _hf0f0f0f0
$constant C: $float<8,23> := _f2.3465e+0
\end{verbatim}
\item Pipe objects:
\begin{verbatim}
$pipe A: $uint<32> $depth 4
$lifo $pipe B: $uint<8> $depth 8
$noblock $pipe C: $uint<39> $depth 2
\end{verbatim}
\item Implicit objects:  these are defined by
statements:
\begin{verbatim}
a := (A + B)
\end{verbatim}
They are also called {\em static-single-assignment}
or SSA variables.
\item Interface objects:  Inputs and outputs of modules are
treated like SSA variables.  Inputs cannot be written into.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Storage Objects}
\begin{itemize}
\item  Will be implemented in memory spaces.
\item  Access to a storage object takes 3+ cycles.
\item  Should use sparingly.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Pipe Objects}
\begin{itemize}
\item Blocking in nature:  provide synchronization mechanism.
\item FIFO or LIFO data transfer within and across modules.
\item Non-blocking pipes: return a 0 if read is attempted on an empty pipe.
\end{itemize}
}
\frame[containsverbatim]{\frametitle{{\bf Aa} SSA Objects}
\begin{itemize}
\item Implemented as registers.
\item Fast access (immediate read, single cycle write).
\item Should use as often as possible.
\item Note that the type of these objects is inferred from the
context in which they appear.
\begin{verbatim}
a := (A + B)
\end{verbatim}
$A$ and $B$ must have the same type, and the type of $a$ is
inferred to be the type of $A$.
\end{itemize}
}

@@

\frame[containsverbatim]{\frametitle{{\bf Aa} Phi Statements}
\begin{itemize}
\item Can appear only in branch blocks, and within merge statements.
\item Indicate the action to be taken depending on how control got
to this statement.
\item  Example:
\begin{verbatim}
       $phi I := ($bitcast($uint<32>) 0) $on $entry
                    NI $on loopback
\end{verbatim}

\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Do-Pipeline-While Statements}
These are not atomic, and can occur only inside a branch-block.
\begin{verbatim}
$dopipeline $depth 7 $buffering 1 $fullrate
   $merge $entry $loopback
        $phi I := ($bitcast ($uint<32>) 0) 
                      $on $entry
                   NI $on $loopback
   $endmerge
   a[I] := (b[I] + c[I])
   NI   := (I+1)
$while (I < 16)
\end{verbatim}
Control-flow:  sequential, controlled by the condition check.  The
places \$entry and \$loopback are implicitly defined.  When control
enters the do-while, the token gets placed in \$entry and when
control loops-back from the condition check, the token gets placed
in \$loopback.  The compiler will pipeline the loop by keeping
\$depth iterations alive.
}

@@

\frame[containsverbatim]{\frametitle{{\bf Aa} Volatile Statements}
\begin{itemize}
\item If you do not want to use a register:
\begin{verbatim}
$volatile p := (q + r)
\end{verbatim}
\item $p$ will not necessarily be implemented as a register.   You
can only assume that $p$ is equivalent to $(q  +  r)$.
\item Useful in describing combinational logic.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{{\bf Aa} Summary}

\begin{itemize}
\item An {\bf Aa} program consists of a collection of module
descriptions and object declarations.
\item Each module is a sequence of statements.
\item Statements can be simple or block-structured.
\item Specific blocks allow branching, while others describe
parallelism.
\item Objects in an {\bf Aa} program can be FIFOs (pipes), storage
or implicitly inferred.
\item Almost everything you can express in C can be expressed, but
we are closer to hardware.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{Look at the FIR again}
}

### 2018-03-05 Documents/Tutorials/Intro/Intro.tex
Updated the tutorial

\documentclass{beamer}
\usepackage{epsfig}
\title{An introduction to the AHIR-v2 tool-set}
\author{Madhav P. Desai}
\begin{document}
\maketitle

@@

\begin{itemize}
\item Aid in the design and implementation of complex systems (in hardware).
\item Describe the algorithm (behaviour) of the system using high-level
languages (currently C), and migrate the algorithm to a hardware
implementation.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{For example}
Design a circuit that takes two inputs, computes the maximum of
the two and returns the maximum.
\begin{verbatim}
uint32_t maxOfTwo(uint32_t a, uint32_t b)
{
  uint32_t c = ((a > b) ? a : b);
}
\end{verbatim}
}

@@

@@

The AHIR-v2 flow translates the code to hardware in four steps:
\begin{enumerate}
\item The source C code is compiled using the open-source compiler
{\bf clang} (we version 2.8), which produces byte-code in {\bf llvm}
format.
\item We take the {\bf llvm} bytecode and translate it to our internal
representation called {\bf Aa}.  This is done by a utility called {\bf llvm2aa}.
\item Multiple {\bf Aa} files can be linked using an {\bf Aa} linker
tool called {\bf AaLinkExtMem}.
\item The linked {\bf Aa} code is then converted to a virtual circuit using
a {\bf vC} representation.  This is done using a utility called {\bf aa2vc}.
\item Finally, the {\bf vC} code is converted to VHDL which can then
be simulated or synthesized.  This conversion is done using a utility
called {\bf vc2vhdl}.
\end{enumerate}
At each stage, various optimizations are carried out.
}

@@

@@

\frame[containsverbatim]{\frametitle{Lets take a look}
Go back to examples/ex1 and check it out.
}

\frame[containsverbatim]{\frametitle{Lets make it a  little more interesting}
\begin{verbatim}
void maxDaemon()
{
  while(1)
  {
       uint32_t a = read_uint32("in_data");
       uint32_t b = read_uint32("in_data");
       c = ((a > b) ? a : b);
       write_uint32("out_data", c);
  }
}
\end{verbatim}
}

\frame[containsverbatim]{\frametitle{The testbench for the Daemon}
\begin{verbatim}
int main(int argc, char* argv[])
{
  uint32_t a, b, c;
  while(1)
  {
     scanf("%d", &a);
     scanf("%d", &b);
     write_uint32("in_data", a);
     write_uint32("in_data", b);
     c = read_uint32("out_data");
     printf(stdout,"Result = %d.\n", c);
  }
  return(0);
}
\end{verbatim}
}

@@

\frame[containsverbatim]{\frametitle{Installing AHIR-v2}

\begin{itemize}
\item You will need to install llvm-2.8, clang-2.8, the BOOST graph library.
\item You can download the source code from gitHub.
\item Go to the v2/ folder in the git repository.  You will need to
source ``build\_bashrc'', and then type ``scons'' to build the executables.
You can then make a release by going back to the root of the git repository
and then typing ``make -f ReleaseMakeFile''.  This populates the Release
directory with binaries, examples etc. which should be enough to get you
going.
\end{itemize}
}

>>>>

\documentclass{beamer}
\usepackage{graphics}
\title{An introduction to the AHIR-v2 tool-set}
\author{Madhav P. Desai\\ Department of Electrical Engg.\\ IIT Bombay, Mumbai India}
\date{March 8, 2018}
\begin{document}
\maketitle

@@

\begin{itemize}
\item Aid in the design and implementation of complex systems (in hardware).
\item Describe the algorithm (behaviour) of the system using high-level
languages, and migrate the algorithm to a hardware
implementation.
\begin{itemize}
\item C.
\item Aa.
\end{itemize}
\end{itemize}
}

\frame[containsverbatim]{\frametitle{Plan}
\begin{itemize}
\item Introduction: going from C to VHDL and simulating the VHDL.
\begin{itemize}
\item Exercise: go through a couple of examples.
\end{itemize}
\item Understand the hardware target model used by the AHIR-V2 tools.
\begin{itemize}
\item Exercise: re-visit the examples to take a closer look.
\end{itemize}
\item Optimizations.
\item Loop-pipelining.
\item {\bf Aa} language.
\item Design problem.
\begin{itemize}
\item Design a programmable FIR filter.
\end{itemize}
\end{itemize}
}

\frame[containsverbatim]{\frametitle{For example}

Design a circuit that takes two inputs, computes the maximum of
the two and returns the maximum.
\begin{verbatim}
uint32_t maxOfTwo(uint32_t a, uint32_t b)
{
  uint32_t c = ((a > b) ? a : b);
}
\end{verbatim}
}

@@

\frame[containsverbatim]{\frametitle{AHIR flow}
\begin{figure}
  \centering
  \includegraphics[width=10cm]{figs/AhirFlow.pdf}
  \caption{AHIR flow}
\end{figure}
}

@@

The AHIR-v2 flow translates the code to hardware in four steps:
\begin{enumerate}
\item The source C code is compiled using the open-source compiler
{\bf clang} (we use version 2.8), which produces byte-code in {\bf llvm}
format.
\item We take the {\bf llvm} bytecode and translate it to our internal
representation called {\bf Aa}.  This is done by a utility called {\bf llvm2aa}.
\item Multiple {\bf Aa} files can be linked using an {\bf Aa} linker
tool called {\bf AaLinkExtMem}.
\item The linked {\bf Aa} code is then converted to a virtual circuit using
a {\bf vC} representation.  This is done using a utility called {\bf aa2vc}.
\item Finally, the {\bf vC} code is converted to VHDL which can then
be simulated or synthesized.  This conversion is done using a utility
called {\bf vc2vhdl}.
\end{enumerate}
At each stage, various optimizations are carried out.
}

@@

\frame[containsverbatim]{\frametitle{VHDL Structure}
\begin{figure}
  \centering
  \includegraphics[width=10cm]{ex1/figs/VhdlStructure.eps}
  \caption{VHDL structure of generated example}
\end{figure}
}

@@

\frame[containsverbatim]{\frametitle{Lets take a look}
Go back to ex1/ and check it out.
}

\frame[containsverbatim]{\frametitle{Lets make it a  little more interesting}
\begin{verbatim}
float x_vals[2];
void firDaemon()
{
  x_vals[0] = 0.0; x_vals[1] = 0.0;
  uint8_t head_pointer = 0;
  // fir loop.
  while(1)
  {
    float x = read_float32("in_data");
    float y = 0.8*x + (0.15*x_vals[head_pointer]) + 
                (0.05*x_vals[(head_pointer+1)&0x1]);
    write_float32("out_data",y);
    head_pointer = (head_pointer + 1)&0x1;
    x_vals[head_pointer] = x;
  }
}
\end{verbatim}
}

\frame[containsverbatim]{\frametitle{The testbench for the Daemon}
\begin{verbatim}
int main(int argc, char* argv[])
{
  while(1)
  {
    float a;
    scanf("%f", &a);
    write_float32("in_data",a);
    float c = read_float32("out_data");
    fprintf(stdout,"Result = %f.\n", c);
  }
  return(0);
}
\end{verbatim}
}

@@

\frame[containsverbatim]{\frametitle{Download and Install}
\begin{itemize}
\item Clone git repository
\begin{verbatim}
git clone https://github.com/madhavPdesai/ahir.git .
\end{verbatim}
\item Install llvm-2.8, clang-2.8 from www.llvm.org (from source, builds on ubuntu 12, centos 6.5, may need  to do minor edits in includes in ubuntu 14).
\item Install boost development packages.
\item Install antlr-2.7.7 (from source, builds on ubuntu 12, centos 6.5, some includes need to be added in ubuntu 14).
\item Install java run-time (openjdk).
\item Install ghdl (or modelsim).
\item Go to release directory. 
\begin{verbatim}
source ahir_bashrc
mkdir bin
./build.sh
\end{verbatim}
\item Should work..
\end{itemize}
}

### 2018-03-05 Documents/Tutorials/Pipelines/Pipelines.tex
Updated the tutorial

\documentclass{beamer}
\usepackage{epsfig}
\title{Describing Pipelines}
\author{Madhav P. Desai}
\begin{document}
\maketitle

\frame[containsverbatim]{\frametitle{Two ways of describing pipelines}

\begin{itemize}
\item with distinct threads.
\item by pipelining a loop.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{An example}
\begin{itemize}
\item We want to compute $d(k) = (a(k)+b(k)) \times c(k)$ for a sequence of 
numbers. 
\item The system should accept a sequence of numbers $\{a(k),b(k),c(k)\}$
and produce the sequence $\{ d(k) \}$.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{As separate threads}
\begin{verbatim}
void Stage_0()
{
  while(1)
  {
    float a = read_float32("a_pipe");
    float b = read_float32("b_pipe");
    float c = read_float32("c_pipe");
    write_float32("c_forward_pipe", c);
    write_float32("s0_result", a+b);
  }
}
void Stage_1()
{
  while(1)
  {
    float c = read_float32("c_forward_pipe");
    float aSb = read_float32("s0_result");
    write_float32("d_pipe", aSb * c);
  }
}
\end{verbatim}
}

\frame[containsverbatim]{\frametitle{In a single loop}
\begin{verbatim}
void Daemon()
{
  while(1)
  {
    float a = read_float32("a_pipe");
    float b = read_float32("b_pipe");
    float c = read_float32("c_pipe");
    write_float32("d_pipe", (a+b)*c);
  }
}
\end{verbatim}
}

\frame[containsverbatim]{\frametitle{Loop Pipelining}
\begin{itemize}
\item Loop pipelining involves executing multiple iterations of a
loop at the same time.
\item While doing this, all dependencies must be taken care of.
\begin{itemize}
\item Operation order.
\item Data dependency.
\item Memory accesses.
\item Pipe accesses.
\end{itemize}
\end{itemize}
}

\frame[containsverbatim]{ \frametitle{Example}
}
\end{document}

>>>>

\documentclass{beamer}
\usepackage{epsfig}
\title{Pipelines:  a fundamental building block for efficient hardware}
\author{Madhav P. Desai\\Department of Electrical Engineering\\IIT-Bombay, Mumbai India}
\date{March 8, 2018}
\begin{document}
\maketitle

\frame[containsverbatim]{\frametitle{An example}
\begin{itemize}
\item We want to compute $result(k) = a(k)*(a(k) + 1.0)$ for a sequence of 
numbers. 
\item The system should accept a sequence of numbers $\{a(k))\}$
and produce the sequence $\{ result(k) \}$.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{The program}
\begin{verbatim}
void Daemon()
{
  while(1)
  {
    float a = read_float32("a_pipe");
    float ap1 = (a + 1.0);
    float result = a*ap1;
    write_float32("result_pipe", result);
  }
}
\end{verbatim}
}

\frame[containsverbatim]{\frametitle{The hardware we would like to obtain}

\begin{figure}
  \centering
  \includegraphics[width=9cm]{figs/ExpectedHardware.pdf}
  \caption{The hardware we would like to see.}
\end{figure}

}

\frame[containsverbatim]{\frametitle{The AHIR  tools produce the expected hardware}
\begin{verbatim}
void __loop_pipelining_on__(int,int,int);
void Daemon()
{
  while(1)
  {
    // tell the AHIR compiler to pipeline this loop..
    // and keep 32 iterations alive at any time.
    __loop_pipelining_on__(32,2,1);

    float a = read_float32("a_pipe");
    float ap1 = (a + 1.0);
    float result = a*ap1;
    write_float32("result_pipe", result);
  }
}
\end{verbatim}

}

\frame[containsverbatim]{\frametitle{Loop Pipelining}
\begin{itemize}
\item The  AHIR compiler modifies the control path to
allow multiple iterations of the loop to be active simultaneously.
\item While doing this, all dependencies must be taken care of.
\begin{itemize}
\item Operation order.
\item Data dependency.
\item Memory accesses.
\item Pipe accesses.
\item Producer-consumer dependencies are the key.
\end{itemize}
\item The data-path is modified by inserting buffers to balance
the pipeline.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{Control path modifications}
\begin{figure}
  \centering
  \includegraphics[width=9cm]{figs/ProducerConsumer.pdf}
  \caption{Producer consumer dependencies in the pipeline}
\end{figure}
}

\frame[containsverbatim]{\frametitle{Cost of loop pipelining}
\begin{itemize}
\item In the data-path: additional buffering.
\item In the control-path, the cost of each join increases by
a factor of $\log N$, where $N$ is the number of iterations that
need to be kept alive.
\item Need to choose $N$ carefully and avoid unbalanced loop bodies
in order to manage cost of loop pipelining.
\item Performance benefits can be substantial.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{An example with and without loop-pipelining}

\begin{itemize}
\item Look at ex1/ and ex1\_no\_loop\_pipeling to see the difference.
\item More than 20X speedup by pipelining the loop.
\end{itemize}
}

\frame[containsverbatim]{\frametitle{Loop-pipelining is critical!}
\begin{itemize}
\item  Think of the hardware as interacting threads, each of which
is a pipelined loop.
\end{itemize}
}

\end{document}
